---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi = 300,
  fig.retina = 2
)
library(tidyverse)
library(ggplot2)
library(ggh4x)
library(purrr)
library(patchwork)
library(cleverly)
```

# cleverly


Cleverly stands for (Cl)ustering with (E)xternal (V)ariabl(e)s in (R) with (L)ongitudinal (Y)s. 


# Installation

You can install the development version of `cleverly` from [GitHub](https://github.com/) with:

```{r load_package, echo = T, results = 'hide', message = F, warning = F}
# install.packages("pak")
pak::pak("empalmer/cleverly")
library(cleverly)
```

> **_NOTE:_**  This package is still in development, so be sure to make sure you have the most up-to-date version. 

Check what version you have installed with:

```{r, echo = T, eval = F}
packageVersion("cleverly") 
```

Then update the package if needed: 

```{r, echo = T, eval = F}
pak::pkg_install("empalmer/cleverly")
```

With package updates, the names and format of the output may change, so be sure to check the documentation and examples. Clustering results should not change, but the format of the output may. 

# Usage: 

> **_NOTE:_** This is a computationally intensive algorithm. Do not be surprised if run time exceeds 10 minutes. Run time can be adjusted by modifying parameters such as `npsi`, `max_admm_iter`, `max_2_iter`.


## Simulate example data 

Simulate compositional count data with longitudinal independence structure. This code will simulate 20 samples across 12 responses, using a binary external variable $Z$. Default arguments to `simulation_data` simulate 3 baseline clusters with 4 members each. 

```{r sim_data}
set.seed(127)
sim <- simulation_data(n = 20,
                       K = 12,
                       user_var = 1000,
                       cor_str = "IND", 
                       Z_type = "binary")

Y <- sim$Y
Z <- sim$Z

true_cluster <- rep(1:3, each = 4)

head(Y)
head(Z)
```

$Y$ is a data frame of 12 counts with a column of time and a column of individual. Z is a vector of binary external variables. 


## Run algorithm: 

We now run the `cleverly` algorithm, specifying the commonly needed arguments. Here, we specify subject_ids = individual, as individual is the id column in Y. We could have alternately given a vector of IDs. Similarly for time. `cleverly` needs to select the ideal tuning parameter `psi` - so we give a range of `psi_min` to `psi_max`, and a number (`npsi`) to test (6). 

We set `cluster_index` to 0 because we want to cluster on the baseline values, i.e. when $Z = 0$. If we want to cluster on the slope value, i.e. the effect of $Z$, we would set `cluster_index` to 1. 

```{r run_algorithm, cache = T}
res <- cleverly(Y = Y,
                Z = Z,
                subject_ids = individual,
                time = time,
                cluster_index = 0,
                cor_str = "IND",
                psi_min = 10,
                psi_max = 5000,
                npsi = 3) 
```

## Results

```{r results}
names(res)
```

`cleverly` outputs include 

 - `clusters`: A list object that contains the cluster membership, number of clusters, and number of responses per cluster
 - `y_hat`: A data frame that includes the fit values, useful for plots and visualizations of the results
 - `BIC`: A list of the BIC values for each psi tested
 - `B`: The B spline basis matrix used for spline fits 
 - `Z`: Copy of the external variable inputted
 - `rho`: Estimated correlation parameter
 - `phi`: Estimated overdispersion parameter
 - `psi`: Chosen hyperparameter
 - `v`: Copy of the pairwise differences 
 - `beta`: Estimated coefficients
 - `possible_clusters`: List of clusters identified for every psi - good for checking if the number and range of psi was sufficient 
 - `error`: NULL unless there was an error in the algorithm. 

## Diagnostics: 

Since we know the true clusters, we can look at the Rand index, adjusted rand index, jaccard index, classification error rate, and number of clusters:


```{r cluster_diag}
get_cluster_diagnostics(res, true_cluster)$cluster_diagnostics
```

We selected the exact right clusters!

## Visualizations

Visualize the data + clusters 

```{r all_plots}
plot_clusters(res)
```

Alternatively, if we want to examine a single cluster: 

```{r one_plot}
plot_one_cluster(res, cluster_val = 1)
```


## Simulate example data (continuous, slope)

`cleverly` also works with continuous external variables, and can additionally cluster on the slope response: 

We first simulation data with a continuous external variable, stored in `sim_cont`, and then run the algorithm.

```{r sim_cont, echo = F}
set.seed(123)
sim_cont <- simulation_data(n = 20,
                        range_start = 5000,
                        range_end = 20000,
                        nknots = 3,
                        K = 12,
                        order = 3,
                        user_var = 1000,
                        cor_str = "CON-d",
                        Z_type = "continuous",
                        rho = 0.4,
                        prob1 = .5,
                        slope_fxns = list(
                          function(t) cos(2 * pi * t),
                          function(t) cos(2 * pi * t),
                          function(t) cos(2 * pi * t),
                          function(t) cos(2 * pi * t),
                          function(t) 2 * sin(pi * t) - 1,
                          function(t) 2 * sin(pi * t) - 1,
                          function(t) 2 * sin(pi * t) - 1,
                          function(t) 2 * sin(pi * t) - 1,
                          function(t) 2 - 2 * t,
                          function(t) 2 - 2 * t,
                          function(t) 2 - 2 * t,
                          function(t) 2 - 2 * t
                        ),
                        # Slope functions
                        baseline_fxns = list(
                          function(t) 2 - t,
                          function(t) 2 * sin(pi * t),
                          function(t) .5,
                          function(t) t^2,
                          function(t)  -.5 ,
                          function(t) 2 * t - 2,
                          function(t) t,
                          function(t) 2 - t,
                          function(t) 1,
                          function(t) -.75,
                          function(t) .75,
                          function(t) 2 * t))
```



```{r run_algorithm_cont, cache = T}
Y_cont <- sim_cont$Y
Z_cont <- sim_cont$Z

head(Z_cont)

res_cont <- cleverly(Y = Y_cont,
                     Z = Z_cont,
                     subject_ids = individual,
                     time = time,
                     cluster_index = 1,
                     cor_str = "IND",
                     psi_min = 10,
                     psi_max = 1000,
                     npsi = 3) 
```

We change `cluster_index` to 1 to indicate that we will cluster based on the slope. 

To plot a slope cluster, we specify `curve_type = "slope"` in the plotting function. We can additionally pass along the response names, as it defaults to 1:nresponses as titles. In the case of a continuous external variable and slope clustering, we also have to pass along the original counts, which are stored in `Y_counts`.

```{r slope_plot}
plot_clusters(res_cont,
              curve_type = "slope",
              response_names = paste0("Response ", 1:12),
              Y_counts = Y_cont[,3:14])
```



## Troubleshooting: 

### Too small hyper parameter range

```{r too_small_psi, cache = T}
res <- cleverly(Y = Y,
                Z = Z,
                subject_ids = individual,
                time = time,
                cluster_index = 0,
                cor_str = "IND",
                psi_min = 10,
                psi_max = 11,
                npsi = 3) 
```

We get a warning, increase the range between `psi_min` and `psi_max`, or increase `npsi`. 


### Poorly fit data

```{r poorly_fit, cache = T}
res <- cleverly(Y = Y,
                Z = Z,
                subject_ids = individual,
                time = time,
                cluster_index = 0,
                cor_str = "IND",
                gammas = c(1e4, 1e4),
                psi_min = 500,
                npsi = 1) 

plot_clusters(res)
```

The tuning parameter `gammas` can be specified, but it is usually fine to leave it at the default. If you see the overall fit seems bad (in this case under fit, because gamma is too large), it can be changed.


# Algorithm details

## Toy example clusters:

```{r toy_example, echo = F, results = 'hide', message = F, warning = F}
baseline_fxns <- list(
  `Taxon 1 - Baseline Cluster A` = function(t) sin(pi * t),
  `Taxon 2 - Baseline Cluster A` = function(t) sin(pi * t),
  `Taxon 3 - Baseline Cluster B` = function(t) cos(2 * pi * t),
  `Taxon 4 - Baseline Cluster B` = function(t) cos(2 * pi * t)
)

slope_fxns <- list(
  function(t) sin(pi * t) + 3 * t,
  function(t) sin(pi * t) + 1,
  function(t) cos(2 * pi * t) + 3 * t,
  function(t) cos(2 * pi * t) + 1
)

# Create data
t_vals <- seq(0, 1, length.out = 100)

baseline_df <- map_dfr(names(baseline_fxns), function(name) {
  tibble(
    t = t_vals,
    value = baseline_fxns[[name]](t_vals),
    function_name = name,
    type = "(Z = 0) - Baseline"
  )
})

slope_df <- map2_dfr(names(baseline_fxns), slope_fxns, function(name, fxn) {
  tibble(
    t = t_vals,
    value = fxn(t_vals),
    function_name = name,
    type = "(Z = 1)"
  )
})

plot_df <- bind_rows(baseline_df, slope_df) %>% 
  mutate(slope_cluster = 
           case_when(
             function_name == "Taxon 1 - Baseline Cluster A" ~ "Taxon 1 - Slope Cluster 1", 
             function_name == "Taxon 2 - Baseline Cluster A" ~ "Taxon 2 - Slope Cluster 2",
             function_name == "Taxon 3 - Baseline Cluster B" ~ "Taxon 3 - Slope Cluster 1",
             function_name == "Taxon 4 - Baseline Cluster B" ~ "Taxon 4 - Slope Cluster 2"
           ))


# Plot
baseline_plot <- ggplot(plot_df, aes(x = t, y = value, color = type, linetype = type)) +
  geom_line(linewidth = 1.5) +
  ggh4x::facet_wrap2(
    ~ function_name,
    strip = strip_themed(
      background_x = elem_list_rect(
        fill = c("#AED9E0","#AED9E0","#90D743",  "#90D743"), 
        color = NA
      ),
      text_x = elem_list_text(
        color = "black"
      )
    )
  ) +
  labs(x = "t", y = "Value", color = "Z", linetype = "Z",
       title = "Cluter on Baseline (solid line)") +
  scale_color_manual(values = c(`(Z = 0) - Baseline` = "#440154",
                                "(Z = 1)" = "#D64F5C")) + 
  theme(
    plot.background = element_rect(color = "black", fill = NA, size = 1)
  )


slope_plot <- plot_df %>% 
  pivot_wider(
    names_from = c(type),
    values_from = value
  ) %>% 
  mutate(diff = `(Z = 1)` - `(Z = 0) - Baseline`) %>%
  ggplot(aes(x = t, y = diff, color = "Slope")) + 
  geom_line(linewidth = 1) + 
  facet_wrap2(
    ~ slope_cluster,
    strip = strip_themed(
      background_x = elem_list_rect(
        fill = c("#D64F5C","#FBB03B", "#D64F5C", "#FBB03B"), 
        color = NA
      ),
      text_x = elem_list_text(
        color = "black"
      )
    )
  ) +
  labs(y = "Difference (Z = 1) - (Z = 0)",
       x = "t", 
       title = "Cluster on slope (differences between lines)" ) + 
  scale_color_manual(name = "",
                     values = c("Slope" = "black")) + 
  theme(
    plot.background = element_rect(color = "black", fill = NA, size = 1)
  )



baseline_plot / slope_plot + 
  plot_annotation(title = "Clustering Taxa using Baseline vs Slope")


```

